---
title: "Dataframes - In Class lab"
author: "Dr. Shoemaker"
date: "DATA 2401"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Working with loaded in csv files


Use the `read.csv()` function to read the data from the `gates_donations.csv` file (available on Canvas, make sure you save it in the project repository folder) into a variable called `grants` 

```{r}
## if you save it in the same folder as this file this should work. 
## But you may need edit the string to reflect the actual file name: 

grants <- read_csv("gates_donations.csv")


head(grants)
```

Because you're in an RProject, you don't need to set your working directory in RStudio! The working directory is automatically set to the repository location. (You can use `getwd()` to confirm this)

- Use the glimpse function to look at the loaded data

```{r}

glimpse(grants)

```


- Create a variable `org` that contains the `organization` column of the dataset

```{r}
data <- read.csv("gates_donations.csv")
glimpse(data)

org <- data$organization
```


- Confirm that the "organization" column is a vector using the `is.vector()` function. (This is a useful debugging tip if you hit errors later!)

```{r}
gates_donations <- read.csv("gates_donations.csv", stringsAsFactors = FALSE)
colnames(gates_donations)
org <- gates_donations$organization  
is.vector(org)



```


Now you can ask some interesting questions about the dataset:

- What was the mean grant value?

```{r}
mean_grant <- mean(grants$total_amount)
print(mean_grant)

```


- What was the dollar amount of the largest grant?

```{r}
largest_grant <- max(grants$total_amount)
print(largest_grant)
```


- What was the dollar amount of the smallest grant?

```{r}
smallest_grant <- min(grants$total_amount)
print(smallest_grant)
```


- Which organization received the largest grant?

```{r}
organization_largest_grant <- max(grants$organization)
print(organization_largest_grant)

```


- Which organization received the smallest grant?

```{r}
organization_smallest_grant <- min(grants$organization)
print(organization_smallest_grant)

grants_2010 <- grants
```

- How many grants were awarded in 2010?

```{r}


grants_2010 <- grants[grants$start_year == 2010, ]  
num_grants_2010 <- nrow(grants_2010)  
print(num_grants_2010)

```

- How much was awarded in 2010? 

```{r}


awards_2010 <- sum(grants$total_amount[grants$start_year == 2010], na.rm = TRUE)
print(awards_2010)

```


# Working with built-in data sets

Load R's "USPersonalExpenditure" dataset using the `data()` function

This will produce a data frame called `USPersonalExpenditure`
```{r}
data("USPersonalExpenditure")
```

The variable `USPersonalExpenditure` is now accessible to you. Unfortunately, it's not a data frame (it's actually a matrix, you can find that out using) 

Test this using the `is.data.frame()` function:

```{r}
is.data.frame(USPersonalExpenditure)
```

Luckily, you can pass the USPersonalExpenditure variable as an argument to the `data.frame()` function to convert it a data frame.

Do this, storing the result in a new variable

```{r}
expenditure_df <- data.frame(USPersonalExpenditure)

```


What are the current column names of your dataframe?
```{r}

colnames(expenditure_df) <- gsub("[^0-9]", "", colnames(expenditure_df))
colnames(expenditure_df) <- as.numeric(colnames(expenditure_df))


print(colnames(expenditure_df))


```


Consider: why are they so strange? (Think about whether you could use a number like 1940 with dollar notation!)

What are the current row names of your dataframe?
```{r}
rownames(expenditure_df)

```


Add a column "Category" to your data frame that contains the rownames

```{r}

expenditure_df$Category <- rownames(expenditure_df)


print(head(expenditure_df))


```

Consider how this data isn't tidy. Use the `pivot_longer()` function from last class to make the data tidy. 

```{r}





```



How much money was spent on personal care in 1940?
```{r}


```


How much money was spent on Food and Tobacco in 1960?
```{r}



```


What was the highest expenditure category in 1960?
```{r}


```


Define a function `lowest_category` that takes in a year as a parameter, and returns the lowest spending category of that year
```{r}



```

```{r}

```


Using your function, determine the lowest spending category of each year
Hint: use the `sapply()` function to apply your function to a vector of years

```{r}

```
